Quantitative Benchmarking for Automatically Generated Conjectures

### 2018-04-30 TEBenchmarkPaper/article.tex
Once-over to make sure we're consistent, etc.

\author{Chris Warburton \and
        Alison Pease    \and
        Jianguo Zhang
        % TODO: Invite Katya? Even if she only offers some help, can acknowledge
}

@@

\emph{Conjecture generation} is the open-ended task of producing conjectures
concerning a given logical theory which are somehow ``interesting'', and is
studied as a sub-field of \emph{mathematical theory exploration} (MTE). This has
applications wherever we find uses for formal methods: in proof assistants and
their libraries, in mathematics education and research, and in the specification
and verification of software. Although formal methods, and automated reasoning
more generally, have been advocated in mathematics (notably by
Voevodsky~\cite{voevodsky2010univalent} and Gowers~\cite{ganesalingam2013fully})
and software engineering (for example via dependently-typed
programming~\cite{McKinna:2006}), they have a reputation for being difficult and
expensive, which has historically limited their use to high-assurance domains
such as aerospace and microprocessor design.

@@

Such use cases require the ability to \emph{discover} true properties of some
particular definitions; this requires not only \emph{proving} their correctness,
but also \emph{posing} them as conjectures in the first place. This is a hard
problem in general, but presents opportunities for automation due to the
precise, symbolic nature of the domain. These processes can also feed into each
other, since proven properties can be used as lemmas for subsequent proofs. As
an example, proving Equation~\ref{eq:mapreduce} may require ``background
knowledge'' which is tedious to produce by hand, such as
$\forall x. \Append(x, \Nil) = x$.

@@

A major challenge when generating conjectures is choosing how to narrow down the
resulting set to only those deemed ``interesting'', since this is an imprecise
term with many different interpretations. For example, all existing approaches
agree that simple tautologies are ``uninteresting'', but differ when it comes to
more complex statements.  Colton \etal{} give a survey of tools for theory
formation and exploration, and their associated notions of
``interestingness'' for concepts and conjectures~\cite{colton2000notion}. Six
notions are identified, which are applied to conjectured properties as follows:

@@

{\bf Utility} is the relevance or usefulness of a property to the user's
particular task. For example, if we want to find optimisations like
Equation~\ref{eq:mapreduce}, then utility would include whether the property
allows rewriting, the difference in resource usage of the expressions involved,
and how common those expressions are in real usage.

@@

The precision and recall of three MTE tools are compared
in~\cite{claessen2013automating}:
IsaCoSy~\cite{Johansson.Dixon.Bundy:conjecture-generation},
IsaScheme~\cite{MontanoRivas2011} and
HipSpec~\cite{Claessen_hipspec:automating}.\footnote{Unfortunately this
  comparison only reports prior results for each tool, rather than reproducing
  them. This makes the numbers less comparable, since each tool was tested with
  slightly different definitions (e.g. whether or not the exponential and
  maximum functions were included).} The definitions and ground truths for all
three were taken from the standard library of the Isabelle proof assistant:
since the library authors have gone to the effort of stating and proving these
properties, and including them with every copy of their software, this is a good
indication that they are interesting. Two sets of definitions were used: a
theory of natural numbers and of lists (which we show in
Figure~\ref{figure:list_theory} on page~\pageref{figure:list_theory}). With only
two datapoints containing only a few definitions and properties (e.g. the
natural number theory contained 4 functions with 12 properties), it is difficult
to draw general conclusions; yet despite these limitations we believe that this
is a viable method to performing meaningful, objective analyses and comparisons
for such tools. It is only with larger data sets that we can judge performance
in different, especially \emph{novel}, domains (which, after all, is the purpose
of MTE tools). In addition, we might be concerned that such ``popular'' theories
as the natural numbers may be unrepresentative of typical usage. For example, we
would not expect the library functions in a typical verification problem to be
as mathematically rich as number theory.

@@

Since our benchmarking methodology builds on the precision/recall analysis
described in $\S$\ref{sec:te}, we decided to use two of the same tools, IsaCoSy
and QuickSpec (the conjecture generating component of HipSpec), as our
demonstration. Other notable tools, including IsaScheme, are described under
related work in $\S$\ref{sec:related-work}.

@@

QuickSpec emerged from work on the QuickCheck software testing framework for the
Haskell programming language~\cite{claessen2011quickcheck}. As well as
generating conjectures for HipSpec (also a Haskell tool) it has also been
applied to Isabelle via the Hipster tool~\cite{Hipster}. Like IsaCoSy, QuickSpec
takes a signature and enumerates well-typed terms, collecting together those of
the same type into equivalence classes, assuming them to be equal. QuickCheck is
then used to find counterexamples to this assumption, which it does by randomly
instantiating the variables and comparing the resulting closed expressions. Any
equivalence class whose elements don't compare equal are split up, and the
process is repeated with new random values.

@@

Some problem sets are more amenable for this purpose than others. The most
suitable are those which have the following properties:

\begin{itemize}
\item For each problem, there should be a clear distinction between the theorem
  to be proved and the definitions involved, such that the two can be easily and
  meaningfully separated. This rules out problem sets like those of
  SMT-COMP~\cite{barrett2005smt}, where many problems involve uninterpreted
  functions, whose behaviour is \emph{implicit} in the logical structure of the
  theorem statement but not separately \emph{defined}.
\item Definitions should be translatable into a form suitable for the MTE tools
  under study. Our application in $\S$\ref{sec:application} requires Haskell and
  Isabelle translations, and also benefits from having definitions be strongly
  typed.
\item The problem set should be relevant to the desired domain. Our focus on
  functional programming requires higher-order functions and inductively defined
  datatypes, which rules out first-order languages/logics (such as
  TPTP~\cite{sutcliffe2009tptp}).
\item The problem set should ideally contain \emph{every} ``interesting''
  property involving its included definitions, since non-membership in the
  ground truth will be treated as being ``uninteresting''. More realistically,
  we should at least aim for each definition to have \emph{multiple} properties;
  rather than a mixture of mostly unrelated properties.
\item Larger problem sets are preferable, all else being equal, as they give
  more robust statistics.
\end{itemize}

Once such a problem set has been chosen, we must separate the definitions
referenced by the theorems from the theorem statements themselves. The former
will be used as input to the tools, whilst the latter form a ground truth corpus
of properties against which to compare the output.

@@

For this reason, duplicates should be removed, and any references in the theorem
statements updated to use the remaining definition (e.g. chosen based on
lexicographic order). In the above example, the \texttt{plus} function would be
removed, and the commutativity statement updated to reference \texttt{add}
instead.

@@

  \end{tikzpicture}
  \caption[]{High-level view of our benchmarking methodology, showing
    \begin{tikzpicture}
      \node [rectangle, text centered, draw=black]{processes};
    \end{tikzpicture},
    \begin{tikzpicture}
      \node [rectangle, rounded corners, text centered, draw=black]{data};
    \end{tikzpicture} and
    \begin{tikzpicture}
      \node [trapezium, trapezium left angle=70, trapezium right angle=110,
             text centered, draw=black]{inputs};
  \end{tikzpicture}}
  \label{figure:flow_chart}
\end{figure}

@@

An important historical implementation of ATF is Lenat's AM system. Unlike prior
work, such as Meta-Dendral~\cite{buchanan:75} and those described
in~\cite{winston}, AM aims to be a general purpose mathematical discovery
system, designed to both construct new concepts and conjecture relationships
between them. AM is a rule-based system which represents knowledge using a
frame-like scheme, enlarges its knowledge base via a collection of heuristic
rules, and controls the firing of these rules via an agenda
mechanism. Evaluation of AM considered generality (performance in new domains)
and how finely-tuned various aspects of the program are (the agenda, the
interaction of the heuristics, etc). Most of this evaluation was qualitative,
and has subsequently been criticised~\cite[chap.~13]{colton:book}. In their
case study in methodology, Ritchie and Hanna found a large discrepancy between
the theoretical claims made of AM and the implemented program~\cite{partridge};
for example, AM ``invented'' natural numbers from sets, but did so using a
heuristic specifically designed to make this connection.

@@

\begin{acknowledgements}
  We are grateful to the authors of the systems we have used in our experiments,
  especially for help in obtaining and configuring their tools. We would
  especially like to thank Koen Claessen, Lucas Dixon, Moa Johansson, Dan
  Ros\'{e}n and Nicholas Smallbone for useful discussions and help in adapting
  their software to our purposes.
%  FIXME: EPSRC, Katya, Jianguo fundung, Alison funding
\end{acknowledgements}

>>>>

\author{Chris Warburton \and
        Alison Pease    \and
        Jianguo Zhang}

@@

\emph{Conjecture generation} is the open-ended task of conjecturing properties
of a given logical theory which are somehow ``interesting'', and is
studied as a sub-field of \emph{mathematical theory exploration} (MTE). This has
applications wherever we find uses for formal methods: in proof assistants and
their libraries, in mathematics education and research, and in the specification
and verification of software. Although formal methods, and automated reasoning
more generally, have been advocated in mathematics (notably by
Voevodsky~\cite{voevodsky2010univalent} and Gowers~\cite{ganesalingam2013fully})
and software engineering (for example via dependently-typed
programming~\cite{McKinna:2006}), they have a reputation for being difficult and
expensive, which has historically limited their use to high-assurance domains
such as aerospace and microprocessor design.

@@

Such use cases require the ability to \emph{discover} true properties of some
particular definitions; this requires not only \emph{proving} their correctness,
but also \emph{posing} them as conjectures in the first place. This is a hard
problem in general, but presents opportunities for automation due to the
precise, symbolic nature of the domain. These processes can also feed into each
other, since proven properties can be used as lemmas for subsequent proofs. As
an example, proving Equation~\ref{eq:mapreduce} may require ``background
knowledge'', such as the property $\forall x. \Append(x, \Nil) = x$, which may
be tedious to produce by hand but possible to discover automatically.

@@

A major challenge when generating conjectures is choosing how to narrow down the
resulting set to only those deemed ``interesting'', since this is an imprecise
term with many different interpretations. For example, all existing approaches
agree that simple tautologies are ``uninteresting'', but differ when it comes to
more complex statements.  Colton \etal{} surveyed tools for theory formation
and exploration, and their associated notions of ``interestingness'' for
concepts and conjectures~\cite{colton2000notion}. Six qualities were identified,
which are applied to conjectured properties as follows:

@@

{\bf Utility} is the relevance or usefulness of a property to the user's
particular task. For example, if we want to find properties which are useful for
optimisation, like Equation~\ref{eq:mapreduce}, then utility would include
whether the property justifies some rewrite rule, the difference in resource
usage of the expressions involved, and how common those expressions are in real
usage.

@@

The precision and recall of three MTE tools are compared
in~\cite{claessen2013automating}:
IsaCoSy~\cite{Johansson.Dixon.Bundy:conjecture-generation},
IsaScheme~\cite{MontanoRivas2011} and
HipSpec~\cite{Claessen_hipspec:automating}.\footnote{Unfortunately this
  comparison only reports prior results for each tool, rather than reproducing
  them. This makes the numbers less comparable, since each tool was tested with
  slightly different definitions (e.g. whether or not the exponential and
  maximum functions were included).} The definitions and ground truths for all
three were taken from the standard library of the Isabelle proof assistant:
since the library authors have gone to the effort of stating and proving these
properties, and including them with every copy of their software, this is a good
indication that they are interesting. Two sets of definitions were used: a
theory of natural numbers; and a theory of lists (which we show in
Figure~\ref{figure:list_theory} on page~\pageref{figure:list_theory}). With only
two datapoints, each containing only a few definitions and properties (4
functions and 12 properties, in the natural number example), it is difficult
to draw general conclusions; yet despite these limitations we believe that this
is a viable method to performing meaningful, objective analyses and comparisons
for such tools. It is only with larger data sets that we can generalise the
measured performance to different, especially \emph{novel}, domains (which,
after all, is the purpose of MTE tools). In addition, we might be concerned that
such ``popular'' theories as the natural numbers may be unrepresentative of
typical usage. For example, we would not expect the library functions in a
typical verification problem to be as mathematically rich as number theory.

@@

Since our benchmarking methodology builds on the precision/recall analysis
described in $\S$\ref{sec:te}, we decided to use two of the same tools, IsaCoSy
and QuickSpec (the conjecture generating component of HipSpec), as our
demonstration.

@@

QuickSpec emerged from work on the QuickCheck software testing framework for the
Haskell programming language~\cite{claessen2011quickcheck}. As well as
conjecturing properties of Haskell definitions for HipSpec, it has been applied
to Isabelle definitions via the Hipster tool~\cite{Hipster}. Like IsaCoSy,
QuickSpec takes a signature and enumerates well-typed terms, collecting together
those of the same type into equivalence classes, assuming them to be
equal. QuickCheck is then used to find counterexamples to this assumption, which
it does by randomly instantiating the variables and comparing the resulting
closed expressions. Any equivalence class whose elements don't compare equal are
split up, and the process is repeated with new random values.

@@

Some problem sets are more amenable for this purpose than others. The most
suitable are those meeting the following criteria:

\begin{itemize}
\item For each problem, there should be a clear distinction between the
  statement to be proved and the definitions involved, such that the two can be
  easily and meaningfully separated. This rules out problem sets like those of
  SMT-COMP~\cite{barrett2005smt}, where many problems involve uninterpreted
  functions, whose behaviour is \emph{implicit} in the logical structure of the
  theorem statement but not separately \emph{defined}.
\item Definitions should be translatable into a form suitable for the MTE tools
  under study. Our application in $\S$\ref{sec:application} requires Haskell and
  Isabelle translations, and also benefits from having definitions be strongly
  typed.
\item The problem set should be relevant to the desired domain. Our focus on
  functional programming requires higher-order functions and inductively defined
  datatypes, which rules out first-order languages/logics (such as
  TPTP~\cite{sutcliffe2009tptp}).
\item The problem set should ideally contain \emph{every} ``interesting''
  property involving its included definitions, since non-membership in the
  ground truth will be treated as being ``uninteresting''. More realistically,
  we should aim for each definition to have \emph{multiple} properties; rather
  than a mixture of unrelated problems, cherry-picked from different domains.
\item Larger problem set are preferable as they give more robust statistics, all
  else being equal (i.e. when this does not sacrifice quality).
\end{itemize}

Once such a problem set has been chosen, we must separate the theorem statements
from the definitions they reference. The latter will be used as input to the
MTE tools, whilst the former constitute the ground truth corpus of properties
against which to compare the output.

@@

For this reason, duplicates should be removed, and any references to them
updated to use the remaining definition (e.g. chosen based on lexicographic
order). In the above example, the \texttt{plus} function would be removed, and
the commutativity statement updated to reference \texttt{add} instead.

@@

  \end{tikzpicture}
  \caption[]{High-level view of the benchmarking methodology described in
    $\S$\ref{sec:proposal}, showing
    \begin{tikzpicture}
      \node [rectangle, text centered, draw=black]{processes};
    \end{tikzpicture},
    \begin{tikzpicture}
      \node [rectangle, rounded corners, text centered, draw=black]{data};
    \end{tikzpicture} and
    \begin{tikzpicture}
      \node [trapezium, trapezium left angle=70, trapezium right angle=110,
             text centered, draw=black]{inputs};
  \end{tikzpicture}}
  \label{figure:flow_chart}
\end{figure}

@@

An important historical implementation of ATF is Lenat's AM (Automated
Mathematician) system. Unlike prior work, such as
Meta-Dendral~\cite{buchanan:75} and those described in~\cite{winston}, AM aims
to be a general purpose mathematical discovery system, designed to both
construct new concepts and conjecture relationships between them. AM is a
rule-based system which represents knowledge using a frame-like scheme, enlarges
its knowledge base via a collection of heuristic rules, and controls the firing
of these rules via an agenda mechanism. Evaluation of AM considered generality
(performance in new domains) and how finely-tuned various aspects of the program
are (the agenda, the interaction of the heuristics, etc). Most of this
evaluation was qualitative, and has subsequently been
criticised~\cite[chap.~13]{colton:book}. In their case study in methodology,
Ritchie and Hanna found a large discrepancy between the theoretical claims made
of AM and the implemented program~\cite{partridge}; for example, AM ``invented''
natural numbers from sets, but did so using a heuristic specifically designed to
make this connection.

@@

\begin{acknowledgements}
  We are grateful to the authors of the systems we have used in our experiments,
  especially for help in obtaining and configuring their tools. We would
  especially like to thank Koen Claessen, Lucas Dixon, Moa Johansson, Dan
  Ros\'{e}n and Nicholas Smallbone for useful discussions and help in adapting
  their software to our purposes.
  (FIXME: Funding details go here)
\end{acknowledgements}
